Table

### 2025-09-15 manuscript/manuscript.qmd
table of terms

---
title: On the spatial aggregation of ecosystem condition indicators
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    orchid: https://orcid.org/0000-0002-9623-9491
    affiliations: 
        - id: nina
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Pb 5685 Torgarden
          city: Trondheim
          postal-code: 7485
    attributes:
        corresponding: true
  - name: Hanno Sandvik
    email: hanno.sandvik@nina.no
    orchid: https://orcid.org/0000-0002-5889-1606
    affiliations:
      - ref: nina
  - name: Bálint Czúcz
    email: balint.czucz@nina.no
    affiliations:
      - ref: nina
  - name: Chloé R. Nater
    email: chloe.nater@nina.no
    affiliations:
      - ref: nina
abstract: |
  Ecosystem integrity is declining world-wide, and the international community is setting targets and deciding on actions to combate the current biodiversity crises. For this to be effective, they need tools for accurately describing the current situation ann to monitor trends over time. Ecosystem condition accounts (ECA) is one such tools that use variables and indicators to describe key ecosystem characteristics, reflecting their condition and deviations from a reference condition. Because the purpose is to inform decisions at relatively high political levels, these metrics are often spatially aggregated to represent larger areas, such as countries. However, spatial aggregation of information has the potential to alter the descriptive and normative interpretations one can make from these metrics. For example, aggregation displacement causes the information held in variables and indicators to diverge when these are aggregated spatially. This process is influenced also by the order of steps involved in normalising and aggregating variables, i.e. the aggregation pathway. Although aggregation displacement and the type of aggregation pathway chosen for the indicator clearly impact both the indicator values and their interpretation, there are no clear guidelines or deliberation on these topics in the SEEA EA standard for ecosystem accounting. This paper outlines the consequences of different aggregation pathways, emphasising their impact on the credibility of ECAs, and how these are interpreted by users. We introduce a standardised terminology for aggregation pathways specific to ecosystem condition indicators following the SEEA EA standard and provide recommendations for selecting appropriate pathways in various contexts. Our discussion of this topic is aimed at raising the general awareness of spatial aggregation issues and to guide indicator developers in choosing and reporting spatial aggregation methods.

>>>>

---
title: On the spatial aggregation of ecosystem condition indicators
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    orchid: https://orcid.org/0000-0002-9623-9491
    affiliations: 
        - id: nina
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Pb 5685 Torgarden
          city: Trondheim
          postal-code: 7485
    attributes:
        corresponding: true
  - name: Hanno Sandvik
    email: hanno.sandvik@nina.no
    orchid: https://orcid.org/0000-0002-5889-1606
    affiliations:
      - ref: nina
  - name: Bálint Czúcz
    email: balint.czucz@nina.no
    affiliations:
      - ref: nina
  - name: Chloé R. Nater
    email: chloe.nater@nina.no
    affiliations:
      - ref: nina
abstract: |
  In face of the ongoing nature crisis, the international community is setting targets and deciding on actions to combate the current biodiversity crises. For this to be effective they need tools to accurately describe the current situation and to monitor trends in ecosystems over time. Ecosystem condition accounts (ECA) is one such tools that use variables and indicators to describe key ecosystem characteristics, reflecting their condition and deviations from a reference condition. Because the purpose is to inform decisions at relatively high political levels, these metrics are often spatially aggregated to represent larger areas, such as countries. However, spatial aggregation of information has the potential to alter the descriptive and normative interpretations one can make from these metrics. For example, aggregation displacement causes the information held in variables and indicators to diverge when these are aggregated spatially. This process is influenced also by the order of steps involved in normalising and aggregating variables, i.e. the aggregation pathway. Although aggregation displacement and the type of aggregation pathway chosen for the indicator clearly impact both the indicator values and their interpretation, there are no clear guidelines or deliberation on these topics in the SEEA EA standard for ecosystem accounting. This paper outlines the consequences of different aggregation pathways, emphasising their impact on the credibility of ECAs, and how these are interpreted by users. We introduce a standardised terminology for aggregation pathways specific to ecosystem condition indicators following the SEEA EA standard and provide recommendations for selecting appropriate pathways in various contexts. Our discussion of this topic is aimed at raising the general awareness of spatial aggregation issues and to guide indicator developers in choosing and reporting spatial aggregation methods.

### 2025-09-15 manuscript/terms_table.qmd
table of terms

```{r tbl-terms}
#| echo: false
terms <- tibble(
  X = as.character(NULL),
  Y = as.character(NULL)
) |>
  add_row(
    X = "Variables, indicators and metrics",
    Y = "We use the SEEA EA definitions of the terms variables, indicators (normalised variables) and metrics (referring to either a variable or an indicator) [@unitednations2024]. Variables are typically either in percentages or fractions, or as raw biophysical units, such as biomass or density. Indicators, on the other hand, are always on a unitless scale between 0 and 1, where 0 represents very poor (or the worst possible) condition, and 1 represents very good (or the best possible) condition. Sometimes indicators are calculated directly [e.g. @åström], but usually indicators are made by normalising variables based on two or more reference levels [@unitednations2024 §5.60]."
  ) |>

@@

    Y = "Reference levels (also called reference values) are variable values that act as anchoring points to specific values on the indicator scale, and are used to convert variables into indicators. This definition is different from that in SEEA EA, but at least conceptually it is the same. There can be from two to many RLs per indicator and they are distinguished with a special nomeclature X~n~ where _n_ is the corresponding indicator value times 100. For example, X~0~ and X~100~ refer to variable values coded as 0 or 1 on the indicator scale, respectively (sometimes called the lower and upper RLs). The X~60~ can hold a special meaning if, like in the WFD and in some applications of SEEA EA, the value 0.6 on the indicator scale represents the boundary or threshold between good and reduced condition. Defining X~60~, or any other RL other than X~0~and X~100~, is one way to quantitatively describe a non-linear relationship between the variable and ecosystem condition (referred to as _Transformation_)."
    Y = "Spatial (sometimes called horizontal) aggregation is the process of synthesising (e.g. averaging) data from different sub-areas and taking this new value as the value for all sub-areas. In related fields of study this operation may be described as _increasing the support_ or _reducing the grain_ [@bierkens2002]. The term _upscaling_ is sometimes used synonymously to our definition of spatial aggregation, but it also has a wider definition tied to the ambiguous use of the terms scale to mean either grain or extent. It's common to lump spatial aggregation together with some types of extra- or interpolation operation, such as averaging point data and attributing this value to entire polygons [@gruijter1990], or spatial estimation methods (e.g. geostatistics) and subsequent spatial sampling [@brus1997]. However, for the detailed deliberation in this study the distinction between these is important to note."
    Y = "Refers to operations involved in combining data from different time points or periods to represenet a single temporal period (e.g. an accounting period or a growing season)."
    X = "Thematic agrgegation",
    Y = "Also called vertical aggregation. In the case of ECA, this refers to an aggregation of two or more indicators into a composite index or indices."

@@

@@

    Y = "Interpolation means to _fill in the gaps_, or to increase the density of the data. Extrapolation means to expand the spatial extent of the data or beyond the current data range. There is often no clear boundary between these two operations when applied in ECA as an area may for example be within the geographical range of the current data, but outside the environmental range. Many different methods are employed to do inter- or extrapolation within ECAs, such as statistical models (e.g. regression models) or geostatistical model (e.g. kriging). These methods are more or less sophisticaled ways to essentially _spread_ or learn from the available data to give values to areas where there is no data. This is distinct from the process and purpose of spatial aggregation, which seeks to simplify data."

@@

    X = "Aggregation dissplacement",
    Y = ""
    Y = ""
    X = "",
    Y = ""

>>>>

```{r tbl-terms}
#| echo: false
terms <- tibble(
  X = as.character(NULL),
  Y = as.character(NULL)
    X = "Variables, indicators and metrics",
    Y = "We use the SEEA EA definitions of the terms variables, indicators (normalised variables) and metrics (referring to either a variable or an indicator) [@unitednations2024]. Variables are typically either in percentages or fractions, or as raw biophysical units, such as biomass or density. Indicators, on the other hand, are always on a unitless scale between 0 and 1, where 0 represents very poor (or the worst possible) condition, and 1 represents very good (or the best possible) condition. Sometimes indicators are calculated directly [e.g. @åström], but usually indicators are made by normalising variables based on two or more reference levels [@unitednations2024 §5.60]."

@@

    Y = "Reference levels (also called reference values) are variable values that act as anchoring points to specific values on the indicator scale, and are used to convert variables into indicators. This definition is different from that in SEEA EA, but at least conceptually it is the same. There can be from two to many RLs per indicator and they are distinguished with a special nomeclature X~n~ where _n_ is the corresponding indicator value times 100. For example, X~0~ and X~100~ refer to variable values coded as 0 or 1 on the indicator scale, respectively (sometimes called the lower and upper RLs). The X~60~ can hold a special meaning if, like in the WFD and in some applications of SEEA EA, the value 0.6 on the indicator scale represents the boundary or threshold between good and reduced condition. Defining X~60~, or any RL other than X~0~ and X~100~, is one way to quantitatively describe a non-linear relationship between the variable and ecosystem condition (referred to as _Transformation_)."

@@

    X = "Aggregation displacement",
    Y = "Referring to the systematic shift that occurs to the relative values of variables and their corresponding indicators when they are spatially aggregated. When we take the average of two variables, each of which have attribute values (i.e. indicator values) associated with them via some non-linear function (e.g. truncation or transformation), then the attribute of the average variable value differs from the average of the attributes of the original variable. @rastetter1992 referred to this as aggregation error and used it to describe some of the dangers associated with upscaling in the sense of using precise fine-scale information to explain or predict large scale phenomena. However, in ECAs, the difference between variables and indicators is an intentional one, and we therefore choose to refer to what is essentially the same mathematical phenomenon with the less derogatory term aggregation displacement."
    Y = "Spatial (sometimes called horizontal) aggregation is the process of summarising data from different sub-areas and taking this new value as the value for all constituent sub-areas. In means going from fine-resolution data to less fine or coarse resolution data. In related fields of study this operation may be described as _increasing the support_ or _reducing the grain_ [@bierkens2002]. Different summary functions can be used, including the area-weighted arithmetic mean, which is required under when aggregation indicator values in SEEA EA. The term _upscaling_ is sometimes used synonymously to our definition of spatial aggregation, but it also has a wider definition tied to the ambiguous use of the terms scale to mean either grain or extent. It's common to lump spatial aggregation together with some types of extra- or interpolation operation, such as averaging point data and attributing this value to entire polygons [@gruijter1990], or spatial estimation methods (e.g. geostatistics) and subsequent spatial sampling [@brus1997]. However, for the detailed deliberation in this study it is important to note the distinction between these."
    Y = "Refers to operations involved in combining data from different time points or periods to represent a single temporal period (e.g. an accounting period or a growing season)."
    X = "Thematic aggregation",
    Y = "Also called vertical aggregation. In the case of ECA, this refers to an aggregation of two or more indicators into a composite index (_plural_: indices)."

@@

    Y = "Interpolation means to _fill in the gaps_, or to increase the density of the data. Extrapolation means to expand the spatial extent of the data or beyond the current data range. There is often no clear boundary between these two operations when applied in ECA as an area may for example be within the geographical range of the current data, but outside the environmental range. Many different methods are employed to do inter- or extrapolation within ECAs, such as statistical models (e.g. regression models) or geostatistical models (e.g. kriging). These methods are more or less sophisticaled ways to essentially _spread_ or learn from the available data to give values to areas where there is no data. This is distinct from the process and purpose of spatial aggregation, which seeks to simplify data."
    Y = "Ecosystem assets in SEEA EA are the smallest spatial units for which information is shought, made up of contiguous spaces of a specific ecosystem type. The ecosystem accounting area is the _'geographical territory for which an ecosystem account is compiled'_ [@unitednations2024]"

@@
